FOODIE, Fortran Object oriented Ordinary Differential Equations integration library based on Abstract Calculus Pattern

### 2015-11-05 papers/announcement/manuscript.tex
improve manuscript

\usepackage{xcolor}
\definecolor{Maroon}{cmyk}{0,0.87,0.68,0.32}
\definecolor{RoyalBlue}{cmyk}{1,0.50,0,0}
\definecolor{gray}{cmyk}{0.01,0.01,0.01,0.01}
\usepackage{listings}
\lstdefinelanguage{MyFortran}[95]{Fortran}{morecomment=[l]{\#},morestring=[d]',morekeywords={procedure,pass,deferred,non_overridable,generic,class,is}}
\lstdefinestyle{code}{%
  basicstyle=\footnotesize,%
  backgroundcolor=\color{gray},%
  language=MyFortran,%
  captionpos=b,%
  columns=fixed,%
  xleftmargin=10pt,%
  numbers=none,%
  numberstyle={\tiny},%
  keywordstyle=\color{RoyalBlue},%
  stringstyle={\sffamily},%
  texcl=true,%
  upquote=true,%
  commentstyle=\color{Maroon}%
}

@@

>>>>

\usepackage{xcolor}
\definecolor{Maroon}{cmyk}{0,0.87,0.68,0.32}
\definecolor{RoyalBlue}{cmyk}{1,0.50,0,0}
\definecolor{gray}{cmyk}{0.01,0.01,0.01,0.01}
\usepackage{listings}
\lstdefinelanguage{MyFortran}[08]{Fortran}{morecomment=[l]{\#},morestring=[d]',morekeywords={procedure,pass,deferred,non_overridable,generic,class,is}}
\lstdefinestyle{code}{%
  basicstyle=\footnotesize,%
  backgroundcolor=\color{gray},%
  language=MyFortran,%
  captionpos=b,%
  columns=fixed,%
  keepspaces=true,%
  xleftmargin=10pt,%
  numbers=none,%
  numberstyle={\tiny},%
  keywordstyle=\color{RoyalBlue},%
  % stringstyle={\sffamily},%
  texcl=true,%
  upquote=true,%
  commentstyle=\color{Maroon}%
}

\lstdefinestyle{codesimple}{%
  basicstyle=\footnotesize,%
  backgroundcolor=\color{gray},%
  captionpos=b,%
  columns=fixed,%
  keepspaces=true,%
  xleftmargin=10pt,%
  numbers=none,%
}

@@

\appendix

\input{src/appendix/euler-1D-parallel-tests-API.tex}

### 2015-11-05 papers/announcement/src/parallel/openmp.tex
improve manuscript

This subsection reports the parallel scaling analysis performed with OpenMP directives-based paradigm. These benchmarks have been done on a \emph{dual Intel(R) Xeon(R) CPU X5650} exacores workstation for a total of 12 physical cores, coupled with 24GB of RAM. In order to perform an accurate analysis 4 different codes have considered:

@@

These codes (see appendix \ref{sec:euler-1D-OpenMP-API} for the implementation details) have been compiled by means of the GNU gfortran compiler v5.2.0 with \emph{-O2 -fopenmp} compilation flags.

The Euler conservation laws are integrated for 10 time steps by means of the TVD RK(5,4) solver: the measured CPU time used for computing the scaling efficiencies is the average of the 10 integrations, thus representing the mean CPU time for computing one time step integration.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.85\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/strong-scaling-comparison.png}
    \caption{Strong scaling, number of cells 100000}\label{fig:strong-scaling-openmp}
  \end{subfigure}\\
  \begin{subfigure}[b]{0.85\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/weak-scaling-comparison.png}
    \caption{Weak scaling, minimum number of cells 10000}\label{fig:weak-scaling-openmp}
  \end{subfigure}\\
  \caption{Scaling efficiency with OpenMP programming model}\label{fig:scaling-openmp}
\end{figure}

For the strong scaling, the benchmark has been conducted with 100000 finite volumes. Figure \ref{fig:strong-scaling-openmp} summarizes the strong scaling analysis: it shows that FOODIE-based code scales similarly to the baseline code without FOODIE.

For the weak scaling the minimum size is 10000 finite volumes and the size is scaled linearly with the OpenMP threads, thus $N_{12} = 120000$ cells. Figure \ref{fig:weak-scaling-openmp} summarizes the weak scaling analysis and it essentially confirms that FOODIE-based code scales similarly to the baseline code without FOODIE.

Both strong and weak scaling analysis point out that for the computing architecture considered the parallel scaling is reasonable up to 8 cores: using 12 cores the measured efficiencies become unsatisfactory, reducing below the 50\%.

@@

\begin{table}[!ht]
  \centering
  \caption{OpenMP benchmarks results\label{tab:openmp-results}}
  \begin{subtable}[b]{0.80\textwidth}
    \centering
    \caption{Strong benchmarks, number of cells 100000\label{tab:openmp-results-strong}}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{ccccc}
      {\sc Number of OpenMP threads} & \multicolumn{4}{c}{\sc CPU time for 1 time step integration}              \\
      \hline
                                     & FOODIE serial & FOODIE parallel & procedural serial & procedural parallel \\
      \cmidrule{2-5}
      1                              & 1.2941        & 1.2942          & 1.2820            & 1.2769              \\
      2                              & /             & 0.7472          & /                 & 0.7784              \\
      4                              & /             & 0.3988          & /                 & 0.4218              \\
      8                              & /             & 0.2185          & /                 & 0.2559              \\
      12                             & /             & 0.2352          & /                 & 0.2272              \\
      \hline
    \end{tabular}}
  \end{subtable}\\
  \begin{subtable}[b]{0.80\textwidth}
    \centering
    \caption{Weak benchmarks, minimum number of cells 10000\label{tab:openmp-results-weak}}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{cccccc}
      {\sc Number of OpenMP threads} & {\sc Number of Cells} & \multicolumn{4}{c}{\sc CPU time for 1 time step integration} \\
      \hline
                                     &                       & FOODIE serial & FOODIE parallel & procedural serial & procedural parallel \\
      \cmidrule{3-6}
      1                              & 10000                 & 0.1496        & 0.1419          & 1.2820            & 1.2769              \\
      2                              & 20000                 & /             & 0.1643          & /                 & 0.7784              \\
      4                              & 40000                 & /             & 0.1671          & /                 & 0.4218              \\
      8                              & 80000                 & /             & 0.1752          & /                 & 0.2559              \\
      12                             & 120000                & /             & 0.2883          & /                 & 0.2272              \\
      \hline
    \end{tabular}}
  \end{subtable}
\end{table}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/strong-cpu-time-comparison.png}
    \caption{Strong benchmark, number of cells 100000}\label{fig:strong-cpu-time-openmp}
  \end{subfigure}\quad%
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/weak-cpu-time-comparison.png}
    \caption{Weak benchmark, minimum number of cells 10000}\label{fig:weak-cpu-time-openmp}
  \end{subfigure}\\
  \caption{CPU time consumed with OpenMP programming model}\label{fig:cpu-time-openmp}
\end{figure}

>>>>

This subsection reports the parallel scaling analysis of Euler 1D test programs (with and without FOODIE) being parallelized by means of OpenMP directives-based paradigm. This parallel model is based on the concept of \emph{threads}: an OpenMP enabled code start a single (master) threaded program and, at run-time, it is able to generate a team of (many) threads that work concurrently on the parallelized parts of the code, thus reducing the CPU time necessary for completing such parts. The parallelization is made by means of \emph{directives} explicitly inserted by the programmer: the communications between threads are automatically handled by the compiler (through the provided OpenMP library used as back-end). OpenMP parallel paradigm is not a standard feature of Fortran, rather it is an extension provided by the compiler vendors. This parallel paradigm constitutes an effective and easy approach for parallelizing legacy serial codes, however its usage is limited to shared memory architectures because all threads must have access to the same memory.

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.80\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/strong-scaling-comparison.png}
    \caption{Strong scaling, number of cells 240000}\label{fig:strong-scaling-openmp}
  \end{subfigure}\\
  \begin{subfigure}[b]{0.80\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/weak-scaling-comparison.png}
    \caption{Weak scaling, minimum number of cells 24000}\label{fig:weak-scaling-openmp}
  \end{subfigure}\\
  \caption{Scaling efficiency with OpenMP programming model}\label{fig:scaling-openmp}
\end{figure}

The benchmarks shown in this section have been done on a \emph{dual Intel(R) Xeon(R) CPU X5650} exacores workstation for a total of 12 physical cores, coupled with 24GB of RAM. In order to perform an accurate analysis 4 different codes have considered:

@@

These codes (see \ref{subsec:euler-1D-OpenMP-API} for the implementation details) have been compiled by means of the GNU gfortran compiler v5.2.0 with \emph{-O2 -fopenmp} compilation flags.

The Euler conservation laws are integrated for 30 time steps by means of the TVD RK(5,4) solver: the measured CPU time used for computing the scaling efficiencies is the average of the 30 integrations, thus representing the mean CPU time for computing one time step integration.

For the strong scaling, the benchmark has been conducted with 240000 finite volumes. Figure \ref{fig:strong-scaling-openmp} summarizes the strong scaling analysis: it shows that FOODIE-based code scales similarly to the baseline code without FOODIE.

For the weak scaling the minimum size is 24000 finite volumes and the size is scaled linearly with the OpenMP threads, thus $N_{12} = 288000$ cells. Figure \ref{fig:weak-scaling-openmp} summarizes the weak scaling analysis and it essentially confirms that FOODIE-based code scales similarly to the baseline code without FOODIE.

Both strong and weak scaling analysis point out that for the computing architecture considered the parallel scaling is reasonable up to 8 cores: using 12 cores the measured efficiencies become unsatisfactory, reducing below the 60\%.

@@

\begin{table}[!ht]
  \centering
  \caption{OpenMP benchmarks results\label{tab:openmp-results}}
  \begin{subtable}[b]{0.80\textwidth}
    \centering
    \caption{Strong benchmarks, number of cells 240000\label{tab:openmp-results-strong}}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{ccccc}
      {\sc Number of OpenMP threads} & \multicolumn{4}{c}{\sc CPU time for 1 time step integration}              \\
      \hline
                                     & FOODIE serial & FOODIE parallel & procedural serial & procedural parallel \\
      \cmidrule{2-5}
      1                              & 3.3466        & 3.3076          & 3.1252            & 3.0873              \\
      2                              & /             & 1.8166          & /                 & 1.7765              \\
      4                              & /             & 0.9798          & /                 & 1.0085              \\
      8                              & /             & 0.5192          & /                 & 0.5055              \\
      12                             & /             & 0.4847          & /                 & 0.4748              \\
      \hline
    \end{tabular}}
  \end{subtable}\\
  \begin{subtable}[b]{0.80\textwidth}
    \centering
    \caption{Weak benchmarks, minimum number of cells 24000\label{tab:openmp-results-weak}}
    \resizebox{1.00\textwidth}{!}{%
    \begin{tabular}{cccccc}
      {\sc Number of OpenMP threads} & {\sc Number of Cells} & \multicolumn{4}{c}{\sc CPU time for 1 time step integration} \\
      \hline
                                     &                       & FOODIE serial & FOODIE parallel & procedural serial & procedural parallel \\
      \cmidrule{3-6}
      1                              & 24000                 & 0.3171        & 0.3162          & 0.3089            & 0.3111              \\
      2                              & 48000                 & /             & 0.3492          & /                 & 0.3854              \\
      4                              & 96000                 & /             & 0.3666          & /                 & 0.4069              \\
      8                              & 192000                & /             & 0.3862          & /                 & 0.4142              \\
      12                             & 288000                & /             & 0.5727          & /                 & 0.6142              \\
      \hline
    \end{tabular}}
  \end{subtable}
\end{table}

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.40\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/strong-cpu-time-comparison.png}
    \caption{Strong benchmark, number of cells 240000}\label{fig:strong-cpu-time-openmp}
  \end{subfigure}\quad%
  \begin{subfigure}[b]{0.40\textwidth}
    \centering
    \includegraphics[width=1.00\textwidth]{openmp_benchmark/euler-1D-openmp/weak-cpu-time-comparison.png}
    \caption{Weak benchmark, minimum number of cells 24000}\label{fig:weak-cpu-time-openmp}
  \end{subfigure}\\
  \caption{CPU time consumed with OpenMP programming model}\label{fig:cpu-time-openmp}
\end{figure}

### 2015-11-05 papers/announcement/src/parallel/parallel.tex
improve manuscript

\begin{itemize}
  \item CoArray Fortran (CAF) model, for shared and distributed memory architectures;
  \item OpenMP directive-based model, for only shared memory architectures;
  \item Message Passing Interface (MPI) library-based model, for only distributed memory architectures;
  \end{itemize}

@@

The same 1D Euler PDE problem is also solved by parallel-enabled codes that are not based on FOODIE: their solutions provide a reference for measuring the effect of FOODIE abstraction on the parallel scaling.

\subsection{CAF benchmark}\label{subsec:caf}

{\color{red} To be written.}

@@

\subsection{MPI benchmark}\label{subsec:mpi}

{\color{red} To be written.}

>>>>

\begin{itemize}
  \item CoArray Fortran (CAF) model, for shared and distributed memory architectures;
  \item OpenMP directive-based model, for only shared memory architectures;
  \end{itemize}

@@

The same 1D Euler PDEs problem is also solved by parallel-enabled codes that are not based on FOODIE: their solutions provide a reference for measuring the effect of FOODIE abstraction on the parallel scaling.

\input{src/parallel/caf.tex}

@@
